---
title: "565 Presentation"
author: "Hayden Atchley, Tristan Parker, Steven Goodsell, Austin Nichalson"
date: "12/8/2021"
output:
  ioslides_presentation:
    transition: faster
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

```{r load_packages, include=FALSE}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(
  char = as.character(read.csv("packages.csv", header = F)[,1])
)
install_github("atchley-sha/R-packageSHA")
library(packageSHA)
```

```{r load_functions, include=FALSE}
source("R/functions.R")
```

## Scenarios

- No build 2040
- 30% move downtown
- 30% move to "neighborhood centers"


## SE Data: No-Build

```{r}
#TAZ by population
#TAZ by jobs
```


## SE Data: Downtown

```{r}
#TAZ by population
#TAZ by jobs
```


## SE Data: Neighborhood Centers

```{r}
#TAZ by population
#TAZ by jobs
```


# Outputs #####################################################


## Mode Choice

```{r}

```


## Congestion: No-Build

```{r load_data_congestion}
noBuild <- read_sf("data/loaded_networks/NoBuild.shp")
DT30 <- read_sf("data/loaded_networks/DT30.shp")
NC30 <- read_sf("data/loaded_networks/NC30.shp")

breaks <- get_breaks(noBuild)$breaks
labels <- get_breaks(noBuild)$labels
```

```{r}
plot_congestion(noBuild, breaks, labels)
```

## Congeestion: Downtown

```{r}
plot_congestion(DT30, breaks, labels)
```

## Congestion: Neighborhood Centers

```{r}
plot_congestion(NC30, breaks, labels)
```

# Congetsion Differences ###############################

```{r}
NBDT <- calculate_differences(new = DT30)
NBNC <- calculate_differences(new = NC30)

NBDTP <- calculate_differences(new = DT30, pct = T)
NBNCP <- calculate_differences(new = NC30, pct = T)
```


## Downtown

```{r}
plot_differences(NBDT, name = "Volume more than No-Build")
```


## Neighborhood Centers

```{r}
plot_differences(NBNC, name = "Volume more than No-Build")
```


## Downtown (percent)

### Non-outliers
```{r}
plot_differences(filter(NBDTP,
                        DIFF >= boxplot.stats(NBDTP$DIFF)$stats[1],
                        DIFF <= boxplot.stats(NBDTP$DIFF)$stats[5]),
                 name = "Times volume more than No-Build")
```

## Downtown (percent)

### Outliers
```{r}
plot_differences(filter(NBDTP,
                        DIFF >= boxplot.stats(NBDTP$DIFF)$stats[5]),
                 name = "Times volume more than No-Build")
```

## Neighborhood Centers (percent)

### Non-outliers
```{r}
plot_differences(filter(NBNCP,
                        DIFF >= boxplot.stats(NBDTP$DIFF)$stats[1],
                        DIFF <= boxplot.stats(NBDTP$DIFF)$stats[5]),
                 name = "Times volume more than No-Build")
```

## Neighborhood Centers (percent)

### Outliers
```{r}
plot_differences(filter(NBNCP,
                        DIFF >= boxplot.stats(NBDTP$DIFF)$stats[5]),
                 name = "Times volume more than No-Build")
```