---
title: "Homework 5: Network Assignmnet and Validation"
author: "Hayden Atchley"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
geometry: margin=1in
output:
  pdf_document:
    latex_engine: lualatex
    includes:
      in_header: "../template/preamble.tex"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

```{r load_packages, include=FALSE}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(
  char = read.csv("../template/packages.csv", header = F)[,1]
  )
p_load_gh("atchley-sha/R-packageSHA")
```

\RaggedRight
\graphicspath{ {./images/} }
\captionsetup[table]{labelformat=empty}

Given a network of

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{network.png}
\end{figure}

and link travel time equations of

```{=latex}
\begin{align*}
t_{AD} &= 20 + 0.01q_{AD}    \\
t_{AC} &= 10 + 0.005q_{AC}   \\
t_{CD} &= 12 + 0.005q_{CD}   \\
t_{BC} &= 7.25 + 0.005q_{BC} \\
t_{BD} &= 20 + 0.01q_{BD},
\end{align*}
```

```{r}
timeConsts <- c(20, 10, 12, 7.25, 20)
timeCoeffs <- c(0.01, 0.005, 0.005, 0.005, 0.01)
totalVols <- c(7000, 5000)
```


with 7000 trips from A to D and 500 from B to D:

\pagebreak

5.1
=====================================

We set up a matrix to solve the system of equations:

```{=latex}
\begin{equation*}
  \begin{blockarray}{ccccccccccc}
  q_{AD} & q_{AC} & q_{CD} & q_{BC} & q_{BD} & t_{AD} & t_{AC} & t_{CD} & t_{BC} & t_{BD} & = \\
  \begin{block}{(cccccccccc|c)}
  0.01   & 0      & 0      & 0      & 0      & -1     & 0      &  0     & 0      &  0     & -20 \topstrut\\
  0      & 0.005  & 0      & 0      & 0      & 0      & -1     &  0     & 0      &  0     & -10 \\
  0      & 0      & 0.005  & 0      & 0      & 0      & 0      &  -1    & 0      &  0     & -12 \\
  0      & 0      & 0      & 0.005  & 0      & 0      & 0      &  0     & - 1    &  0     & -7.25 \\
  0      & 0      & 0      & 0      & 0.01   & 0      & 0      &  0     & 0      &  -1    & -20\\
  0      & 0      & 0      & 0      & 0      & -1     & 1      &  1     & 0      &  0     & 0 \\
  0      & 0      & 0      & 0      & 0      & 0      & 0      &  1     & 1      &  -1    & 0 \\
  1      & 1      & 0      & 0      & 0      & 0      & 0      &  0     & 0      &  0     & 7000 \\
  0      & 0      & 0      & 1      & 1      & 0      & 0      &  0     & 0      &  0     & 5000 \\
  0      & 1      & -1     & 1      & 0      & 0      & 0      &  0     & 0      &  0     & 0 \bottomstrut\\
  \end{block}
  \end{blockarray}
\end{equation*}
```

The first about half of these rows come from the trip time equations. The remaining rows come from the following equations:
```{=latex}
\begin{align*}
 t_{AD} &= t_{AC} + t_{CD} \\
 t_{BD} &= t_{BC} + t_{CD} \\
 q_{AD} + q_{AC} &= 7000 \\
 q_{BD} + q_{BC} &= 5000 \\
 q_{CD} &= q_{AC} + q_{BC}.
\end{align*}
```


Solving this matrix gives us:
```{r}
UE <- matrix(c(0.01,0,0,0,0,-1,0,0,0,0,-20,
0,0.005,0,0,0,0,-1,0,0,0,-10,
0,0,0.005,0,0,0,0,-1,0,0,-12,
0,0,0,0.005,0,0,0,0,-1,0,-7.25,
0,0,0,0,0.01,0,0,0,0,-1,-20,
0,0,0,0,0,-1,1,1,0,0,0,
0,0,0,0,0,0,0,1,1,-1,0,
1,1,0,0,0,0,0,0,0,0,totalVols[1],
0,0,0,1,1,0,0,0,0,0,totalVols[2],
0,1,-1,1,0,0,0,0,0,0,0
         ),
       byrow = T, ncol = 11) %>%
  rref()
UE %>% kbl()
```

\medskip

or:
```{r}
UEAns <- tibble(
  "$q_{AD}$" = numeric(),
  "$q_{AC}$" = numeric(),
  "$q_{CD}$" = numeric(),
  "$q_{BC}$" = numeric(),
  "$q_{BD}$" = numeric(),
  "$t_{AD}$" = numeric(),
  "$t_{AC}$" = numeric(),
  "$t_{CD}$" = numeric(),
  "$t_{BC}$" = numeric(),
  "$t_{BD}$" = numeric()
)
UEAns[1,] <- as.list(UE[,11])
UEAns %>%
  kbl(booktabs = T,
      position = "h",
      linesep = "",
      escape = F
      ) %>%
  column_spec(5, border_right = T) %>%
  kable_styling(
    latex_options = c("striped", "HOLD_position", "repeat_header"),
    full_width = F
    )
```

\pagebreak

5.2
====================================

```{r create0table}
iter0 <- tibble(
  "$n$" = character(),
  "$q_{AD}$" = numeric(),
  "$q_{AC}$" = numeric(),
  "$q_{CD}$" = numeric(),
  "$q_{BC}$" = numeric(),
  "$q_{BD}$" = numeric(),
  "$t_{AD}$" = numeric(),
  "$t_{AC}$" = numeric(),
  "$t_{CD}$" = numeric(),
  "$t_{BC}$" = numeric(),
  "$t_{BD}$" = numeric(),
  "$T_{AD}$" = numeric(),
  "$T_{ACD}$" = numeric(),
  "$T_{BD}$" = numeric(),
  "$T_{BCD}$" = numeric()
  )

iter0[1,] <- c(list(as.character(0),0,0,0,0,0),
               timeConsts,
               list(timeConsts[1],
                    sum(timeConsts[2:3]),
                    timeConsts[5],
                    sum(timeConsts[4:3])
                    )
              )
```

```{r}
read_csv("data/5-2_AON.csv") %>%
  kbl(booktabs = T,
      linesep = "",
      escape = F,
      caption = "All-or-nothing (AON) Averages"
      ) %>%
  column_spec(c(1,6,11,13), border_right = T) %>%
  row_spec(4, hline_after = T) %>%
  kable_styling(
    latex_options = c("striped", "HOLD_position", "scale_down", "repeat_header")
    )
```


5.3
===================================

```{r}
read_csv("data/5-3_INCR.csv") %>%
  kbl(booktabs = T,
      linesep = "",
      escape = F,
      caption = "Incremental Assignment"
      ) %>%
  column_spec(c(1,6,11,13), border_right = T) %>%
  row_spec(4, hline_after = T) %>%
  kable_styling(
    latex_options = c("striped", "HOLD_position", "scale_down", "repeat_header")
    )
```


5.4
===================================

```{r}
n = 5

fhwa <- iter0
f <- iter0[,2:6]

for(i in 1:(n+1)){
  phi <- 1/i
  
  AVol <- if(fhwa[i,]$`$T_{AD}$` < fhwa[i,]$`$T_{ACD}$`) 1 else c(2,3)
  BVol <- if(fhwa[i,]$`$T_{BD}$` < fhwa[i,]$`$T_{BCD}$`) 5 else c(4,3)
  
  f[i,] <- 0
  f[i,AVol] <- f[i,AVol] + totalVols[1]
  f[i,BVol] <- f[i,BVol] + totalVols[2]
  
  fhwa[i+1,1] = as.character(i)
  fhwa[i+1,2:6] =
    (1-phi)*fhwa[i,2:6] + phi*f[i,]
  
  incrVols <- fhwa[i+1,2:6] %>% as_vector() %>% unname
  
  fhwa[i+1,7:11] = 
    (timeConsts + timeCoeffs * incrVols) %>%
    as.list()
  fhwa[i+1,12:15] = 
    as.list(c(
      fhwa[i+1,7],
      sum(fhwa[i+1,8:9]),
      fhwa[i+1,11],
      sum(fhwa[i+1,10:9])
    ))
  
}

fhwa[n+2,] <- fhwa[n+1,]
fhwa[n+2,1] <- "Final"

fhwa %<>% as.data.frame()
fhwa[,2:6] %<>% format(digits = 0)
fhwa[,7:15] %<>% format(digits = 1)

fhwa %>%
  kbl(booktabs = T,
      linesep = "",
      escape = F,
      caption = "FHWA Successive Averages",
      digits = 2
      ) %>%
  kable_styling(
    latex_options = c("striped", "HOLD_position", "scale_down", "repeat_header")
    ) %>%
  column_spec(c(1,6,11,13), border_right = T) %>%
  row_spec(n+1, hline_after = T)


```


























